### 2024-12-24 2024-5-13/sections/implementation.tex
complete complexity analysis

In total, the NestC benchmark suite contains 1000 instances.

@@

We evaluate the performance of $\ostrichrecl$ against the state-of-the-art string constraint solvers, including CVC5
\cite{cvc5}, Z3seq \cite{z3seq}, Z3str3
\cite{Z3-str3}, Z3str3RE \cite{BD+23}, and OSTRICH
\cite{CHL+19}, on RegCoL, AutomatArk and NestC benchmark suites. 
The experiment results can be found in Table~\ref{tab:table_overall_eval}. Note that we take the results of CVC5 as the ground truth\footnote{Initially,  we used the majority vote of the results of the solvers as the ground truth. Nevertheless, on some problem instances, all the results of the three solvers in the Z3 family are wrong (after manual inspection), thus failing this approach on these instances.}, and the results different from the ground truth are classified as \emph{error}. We can see that $\ostrichrecl$ solves almost all instances, except 182 of them, that is, it solves 49,638 instances correctly. The number is 4,547/1,917/13,618/2,764/3,134 more than the number of instances solved by CVC5/Z3str3RE/Z3str3/Z3seq/OSTRICH respectively.
%
Moreover, $\ostrichrecl$ is the second fastest solver, whose average time on each instance is close to the fastest solver Z3str3RE (2.88s versus 1.99s).

>>>>

In total, the NestC benchmark suite contains 1000 \hdh{tough} instances, \hdh{where the regexes and their associated counting bounds are generated randomly.}

@@

We evaluate the performance of $\ostrichrecl$ against the state-of-the-art string constraint solvers, including CVC5~\cite{cvc5}, Z3seq~\cite{z3seq}, Z3str3 \cite{Z3-str3}, Z3str3RE~\cite{BD+23}, and OSTRICH \cite{CHL+19}, on RegCoL, AutomatArk and NestC benchmark suites. 
The experiment results can be found in Table~\ref{tab:table_overall_eval}. Note that we take the results of CVC5 as the ground truth\footnote{Initially,  we used the majority vote of the results of the solvers as the ground truth. Nevertheless, on some problem instances, all the results of the three solvers in the Z3 family are wrong (after manual inspection), thus failing this approach on these instances.}, and the results different from the ground truth are classified as \emph{error}. We can see that $\ostrichrecl$ solves almost all instances, except 182 of them, that is, it solves 49,638 instances correctly. The number is 4,547/1,917/13,618/2,764/3,134 more than the number of instances solved by CVC5/Z3str3RE/Z3str3/Z3seq/OSTRICH respectively.
%
Moreover, $\ostrichrecl$ is the second fastest solver, whose average time on each instance is close to the fastest solver Z3str3RE (2.88s versus 1.99s).

### 2024-12-24 2024-5-13/sections/introduction.tex
complete complexity analysis

\sout{Nevertheless, }The aforementioned state-of-the-art string constraint solvers still suffer from such string constraints, especially when the counting and length bounds are large. For instance, none of the string solvers CVC5, Z3seq, Z3-Trau, Z3str3, Z3str3RE, and OSTRICH is capable of solving the following constraint within 120 seconds,
%, and Z3 can not solve it within 120 seconds.
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three strings $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur (called RECL for brevity). We make the following contributions.
\begin{itemize}
  \item We propose an automata-theoretical approach for solving RECL constraints.
 Our main idea is to represent the counting operators by cost registers in cost-enriched finite automata (CEFA, see Section~\ref{sec:automaton} for the definition), instead of unfolding them explicitly. The string-length function is modeled by cost registers as well. The satisfiability of RECL constraints is reduced to the non-emptiness problem of CEFA w.r.t. a linear integer arithmetic (LIA) formula. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA.
Thus, the non-emptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers. Moreover, we propose techniques to reduce the sizes (i.e., the number of states and transitions) of CEFA in order to achieve better performance.
%
  \item For complex RECL constraints where the counting operators may be nested with the other counting operators or complement operators, we propose optimization techniques to improve the performance. Furthermore, for the satisfiable RECL constraints where the solutions of the LIA formulas can be obtained, but the models of the original RECL constraint cannot be generated,  we propose optimization techniques to accelerate the model generation process. 
%
%  \item 
%  Combined with the size-reduction techniques mentioned above, the register representation of regex-counting and string-length in CEFA entails an efficient algorithm for solving RECL constraints. 
 %
\item We implement the algorithm on top of OSTRICH, resulting in a string solver called $\ostrichrecl$\footnote{The tool is available as open source at \url{https://github.com/SimpleXiaohu/ostrich/tree/jsa2024}.}. Finally, we utilize a collection of benchmark suites comprising 49,843 problem instances\footnote{The benchmark suites are available at \url{https://github.com/SimpleXiaohu/ostrich/tree/jsa2024/zaligvinder/models/jsa2024}.} in total to evaluate the performance of $\ostrichrecl$. The experiment results show that $\ostrichrecl$ solves the RECL constraints more efficiently than the state-of-the-art string solvers (see Table~\ref{tab:table_overall_eval}-\ref{tab:results_model-gen}). For instance, $\ostrichrecl$ solves 49,638 out of 49,843 instances, while the best state-of-the-art solver solve only 47,721 instances. We also do experiments to justify some of our technical choices and evaluate the performance of the optimizations. 
%
% 1,000 benchmark instances where the counting and length bounds are large or the counting operators are nested with some other counting operators or complement operators, $\ostrichrecl$ solves at least $278$ more instances than the other solvers, while spending only half or less time per instance on average. 
\end{itemize}

\medskip
\noindent
\emph{Related work.} 
We discuss more related work on regexes with counting, string-length function, and automata with registers/counters.  
%
Determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. 
%
Real-world regexes in programming languages include features beyond classical regexes, e.g., the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. 
Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22}. 
%
The Trau tool focuses on string constraints involving flat regular languages and string-length function and solves them by computing LIA formulas that define the Parikh images of flat regular languages~\cite{z3trau}. The Slent tool solves the string constraints involving string-length function by encoding them into so-called length-encoded automata, then utilizing symbolic model checkers to solve their non-emptiness problem~\cite{WC+18}. However, neither Trau nor Slent supports counting operators explicitly, in other words, counting operators in regexes should be unfolded before solved by them.
%
Cost registers in CEFAs are different from registers in (symbolic) register automata~\cite{ra,sra}:  In register automata, registers are used to store input values and can only be compared for equality/inequality, while in CEFAs, cost registers are used to store integer-valued costs and can be updated by adding/subtracting integer constants and constrained by the accepting conditions which are LIA formulas.
%
Therefore, cost registers in CEFAs are more like counters in counter automata/machines~\cite{Minsky67}, that is, CEFAs can be obtained from counter machines by removing transition guards and adding accepting conditions. \hdh{CEFAs can also be considered as Parikh automata \cite{parikh_automaton_2003} equipped with the ability to decrease counters.}
%
%For the automata model, the definition of registers of CEFAs resembles the counters of the counter automata class~\cite{Minsky67}, but CEFAs contain no guards in each transition. 
%
Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to quickly match a subclass of regexes with counting. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the non-emptiness problem of these automata was not considered, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.

>>>>

\sout{Nevertheless, }The aforementioned state-of-the-art string constraint solvers still suffer from such string constraints, especially when the counting and length bounds are large. For instance, none of the string solvers CVC5, Z3seq, Z3-Trau, Z3str3, Z3str3RE, and OSTRICH is capable of solving the following constraint within 120 seconds,
%, and Z3 can not solve it within 120 seconds.
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three strings $x_1, x_2, x_3$ where $a$ (resp. $b$, $c$) does not occur in $x_1$ (resp. $x_2$, $x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur (called RECL for brevity). We make the following contributions.
\begin{itemize}
  \item We propose an automata-theoretical approach for solving RECL constraints.
 Our main idea is to represent the counting operators by cost registers in cost-enriched finite automata (CEFA, see Section~\ref{sec:automaton} for the definition), instead of unfolding them explicitly. The string-length function is modeled by cost registers as well. The satisfiability of RECL constraints is reduced to the non-emptiness problem of CEFA w.r.t. a linear integer arithmetic (LIA) formula. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA.
Thus, the non-emptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers. Moreover, we propose techniques to reduce the sizes (i.e., the number of states and transitions) of CEFA in order to achieve better performance.
%
  \item For complex RECL constraints where the counting operators may be nested with the other counting operators or complement operators, we propose optimization techniques to improve the performance. Furthermore, for the satisfiable RECL constraints where the solutions of the LIA formulas can be obtained, but the models of the original RECL constraint cannot be generated,  we propose optimization techniques to accelerate the model generation process. 
%
%  \item 
%  Combined with the size-reduction techniques mentioned above, the register representation of regex-counting and string-length in CEFA entails an efficient algorithm for solving RECL constraints. 
 %
\item We implement the algorithm on top of OSTRICH, resulting in a string solver called $\ostrichrecl$. Finally, we utilize a collection of benchmark suites comprising 49,843 problem instances in total to evaluate the performance of $\ostrichrecl$. The experiment results show that $\ostrichrecl$ solves the RECL constraints more efficiently than the state-of-the-art string solvers (see Table~\ref{tab:table_overall_eval}-\ref{tab:results_model-gen}). For instance, $\ostrichrecl$ solves 49,638 out of 49,843 instances, while the best state-of-the-art solver solve only 47,721 instances. We also do experiments to justify some of our technical choices and evaluate the performance of the optimizations\footnote{\hdh{The tool is available as open source at \url{https://github.com/SimpleXiaohu/ostrich/tree/jsa2024}.

The benchmark suites are available at \url{https://github.com/SimpleXiaohu/ostrich/tree/jsa2024/zaligvinder/models/jsa2024}.}}. 
%
% 1,000 benchmark instances where the counting and length bounds are large or the counting operators are nested with some other counting operators or complement operators, $\ostrichrecl$ solves at least $278$ more instances than the other solvers, while spending only half or less time per instance on average. 
\end{itemize}

\medskip
\noindent
\emph{Related work.} 
We discuss more related work on regexes with counting, string-length function, and automata with registers/counters.  
%
Determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. 
%
Real-world regexes in programming languages include features beyond classical regexes, e.g., the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. 
Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22}. 
%
The Trau tool focuses on string constraints involving flat regular languages and string-length function and solves them by computing LIA formulas that define the Parikh images of flat regular languages~\cite{z3trau}. The Slent tool solves the string constraints involving string-length function by encoding them into so-called length-encoded automata, then utilizing symbolic model checkers to solve their non-emptiness problem~\cite{WC+18}. However, neither Trau nor Slent supports counting operators explicitly, in other words, counting operators in regexes should be unfolded before solved by them.
%
Cost registers in CEFAs are different from registers in (symbolic) register automata~\cite{ra,sra}:  In register automata, registers are used to store input values and can only be compared for equality/inequality, while in CEFAs, cost registers are used to store integer-valued costs and can be updated by adding/subtracting integer constants and constrained by the accepting conditions which are LIA formulas.
%
Therefore, cost registers in CEFAs are more like counters in counter automata/machines~\cite{Minsky67}, that is, CEFAs can be obtained from counter machines by removing transition guards and adding accepting conditions. \hdh{CEFAs can also be considered as a variation of Parikh automata \cite{parikh_automaton_2003} in which the numbers in the alphabet can be integers, instead of natural numbers.}
%
%For the automata model, the definition of registers of CEFAs resembles the counters of the counter automata class~\cite{Minsky67}, but CEFAs contain no guards in each transition. 
%
Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to quickly match a subclass of regexes with counting. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the non-emptiness problem of these automata was not considered, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.

### 2024-12-24 2024-5-13/sections/overview.tex
complete complexity analysis

\begin{figure}[ht]
\vspace{-2mm}
  \centering
  \includegraphics[width = 0.8\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cB$ and $\cC$}
  \label{fig:overview:product:reduced}
\vspace{-4mm}
\end{figure}
%
Let $\costset = \{\myvec{v_1}, \cdots, \myvec{v_m}\}$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi_\cB(\anivar_1, \cdots, \anivar_m)$ can be computed to define the Parikh image of strings that are accepted by $\cB$, where $\anivar_1, \cdots, \anivar_m$ are the integer variables to denote the number of occurrences of $\myvec{v_1}, \cdots, \myvec{v_m}$.
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula,
\begin{equation}\label{eqn-LIA}
\begin{array}{l}
\psi_\cB(\anivar_1, \cdots, \anivar_m) \wedge \bigwedge \limits_{1 \le j \le 4} r_j = \sum \limits_{1\le k \le m}  \anivar_k v_{k, j}\ \wedge \\
1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_4 > 120.
\end{array}
\end{equation}
which can be solved by the off-the-shelf SMT solvers.

>>>>

\begin{figure}[ht]
\vspace{-2mm}
  \centering
  \includegraphics[width = 0.8\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cB$ and $\cC$}
  \label{fig:overview:product:reduced}
\vspace{-4mm}
\end{figure}
%
Let $\costset = \{\myvec{v}_1, \cdots, \myvec{v}_m\}$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi_\cB(\anivar_1, \cdots, \anivar_m)$ can be computed to define the Parikh image of strings that are accepted by $\cB$, where $\anivar_1, \cdots, \anivar_m$ are the integer variables to denote the number of occurrences of $\myvec{v}_1, \cdots, \myvec{v}_m$.
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula,
\begin{equation}\label{eqn-LIA}
\begin{array}{l}
\psi_\cB(\anivar_1, \cdots, \anivar_m) \wedge \bigwedge \limits_{1 \le j \le 4} r_j = \sum \limits_{1\le k \le m}  \anivar_k v_{k, j}\ \wedge \\
1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_4 > 120.
\end{array}
\end{equation}
which can be solved by the off-the-shelf SMT solvers.
